---
title: "count composition"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
---


```{r libraries, eval=TRUE, include=FALSE}
library(targets)
library(compositions)
library(gridExtra)
library(ggplot2)

source("scripts/helper_functions.R")

D <- 13
basis_vectors <- lapply(1:(D - 1), generate_orthonormal_basis, D)
basis_matrix <- do.call(rbind, basis_vectors)
```

The MCEM Algorith identified the following principal components for the whole dataset:

```{r load data, eval=TRUE}
pca_count <- tar_read(pca_count_ilr_vs1_1perc)
pca <- pca_count$pca
count_data <- pca_count$x_data
count_matrix <- do.call(rbind, count_data)
x_count <- tar_read(data_kl15)
x_clr <- tar_read(data_kl15_comp_clr)

count_acomp <- acomp(count_matrix)

# pc is needs to be transformed to the clr space
pca_rotation_clr <- t(basis_matrix) %*% pca$rotation %*% basis_matrix
colnames(count_matrix) <- gsub("_cts$", "_clr", colnames(count_matrix))
rownames(pca_rotation_clr) <- colnames(count_matrix)

# first four components
pca_rotation_clr[, 1:4]
```

## Screeplot

```{r screeplot, eval=TRUE}
plot(pca)

sum(pca$sdev[1:2]^2)/sum(pca$sdev^2)
```

We can calculate the percantage of explained variance by the principal components with the **variation matrix**:
The first two components explain `r sum(pca$sdev[1:2]^2)/sum(pca$sdev^2) * 100` % of the total variance.  
The first four components explain `r sum(pca$sdev[1:4]^2)/sum(pca$sdev^2) * 100` % of the total variance.  
The first component explains `r pca$sdev[1]^2/sum(pca$sdev^2) * 100` % of the total variance.  
The second component explains `r pca$sdev[2]^2/sum(pca$sdev^2) * 100` % of the total variance.  
The third component explains `r pca$sdev[3]^2/sum(pca$sdev^2) * 100` % of the total variance.  
The fourth component explains `r pca$sdev[4]^2/sum(pca$sdev^2) * 100` % of the total variance.  

## Biplot

```{r biplot, eval=TRUE}
plot_pca_rotations(pca_rotation_clr, components = c(1,2), main = "PC1 vs. PC2")
plot_pca_rotations(pca_rotation_clr, components = c(1,3), main = "PC1 vs. PC3")
plot_pca_rotations(pca_rotation_clr, components = c(1,4), main = "PC1 vs. PC4")


```

## Principal Components

```{r principal components, eval=TRUE, fig.width=12, fig.height=5}
center_clr <- t(basis_matrix) %*% pca$center

# Create data frame for plotting
mean_comp_1 <- data.frame(
    component = 1:13,
    mean_value = center_clr,
    pc1_plus = center_clr + pca_rotation_clr[,1] * pca$sdev[1],
    pc1_minus = center_clr - pca_rotation_clr[,1] * pca$sdev[1]
)

mean_comp_2 <- data.frame(
    component = 1:13,
    mean_value = center_clr,
    pc1_plus = center_clr + pca_rotation_clr[,2] * pca$sdev[2],
    pc1_minus = center_clr - pca_rotation_clr[,2] * pca$sdev[2]
)

mean_comp_3 <- data.frame(
    component = 1:13,
    mean_value = center_clr,
    pc1_plus = center_clr + pca_rotation_clr[,3] * pca$sdev[3],
    pc1_minus = center_clr - pca_rotation_clr[,3] * pca$sdev[3]
)

mean_comp_4 <- data.frame(
    component = 1:13,
    mean_value = center_clr,
    pc1_plus = center_clr + pca_rotation_clr[,4] * pca$sdev[4],
    pc1_minus = center_clr - pca_rotation_clr[,4] * pca$sdev[4]
)


# Add component names if available
mean_comp_1$names <- colnames(count_matrix)
mean_comp_2$names <- colnames(count_matrix)
mean_comp_3$names <- colnames(count_matrix)
mean_comp_4$names <- colnames(count_matrix)

plot1 <- ggplot(mean_comp_1) +
    geom_bar(aes(x = names, y = pc1_plus, color = "Factor 0.35"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = pc1_minus, color = "Factor -0.35"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = mean_value), 
             stat = "identity", fill = NA, color = "black") +
    scale_color_manual(values = c("Factor 0.35" = "red", "Factor -0.35" = "blue"),
                      name = "PC1",
                      breaks = c("Factor 0.35", "Factor -0.35")) +
    labs(x = "Component", y = "clr value",
         title = "Mean composition with PC1 variations") +
    theme_grey() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "top")


plot2 <- ggplot(mean_comp_2) +
    geom_bar(aes(x = names, y = pc1_plus, color = "Factor 0.23"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = pc1_minus, color = "Factor -0.23"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = mean_value), 
             stat = "identity", fill = NA, color = "black") +
    scale_color_manual(values = c("Factor 0.23" = "red", "Factor -0.23" = "blue"),
                      name = "PC2",
                      breaks = c("Factor 0.23", "Factor -0.23")) +
    labs(x = "Component", y = "clr value",
         title = "Mean composition with PC2 variations") +
    theme_grey() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "top")


plot3 <- ggplot(mean_comp_3) +
    geom_bar(aes(x = names, y = pc1_plus, color = "Factor 0.11"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = pc1_minus, color = "Factor -0.11"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = mean_value), 
             stat = "identity", fill = NA, color = "black") +
    scale_color_manual(values = c("Factor 0.11" = "red", "Factor -0.11" = "blue"),
                      name = "PC3",
                      breaks = c("Factor 0.11", "Factor -0.11")) +
    labs(x = "Component", y = "clr value",
         title = "Mean composition with PC3 variations") +
    theme_grey() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "top")


plot4 <- ggplot(mean_comp_4) +
    geom_bar(aes(x = names, y = pc1_plus, color = "Factor 0.09"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = pc1_minus, color = "Factor -0.09"), 
             stat = "identity", fill = NA) +
    geom_bar(aes(x = names, y = mean_value), 
             stat = "identity", fill = NA, color = "black") +
    scale_color_manual(values = c("Factor 0.09" = "red", "Factor -0.35" = "blue"),
                      name = "PC4",
                      breaks = c("Factor 0.09", "Factor -0.09")) +
    labs(x = "Component", y = "clr value",
         title = "Mean composition with PC4 variations") +
    theme_grey() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "top")

grid.arrange(plot1,plot2,plot3,plot4, ncol = 4)
```

## Scores 

The scores for the principal component $k$ can be easily computed with 
$\boldsymbol{z}_i = \boldsymbol{\varphi}_k^{\top} (\operatorname{clr}(\boldsymbol{x}_i) - \boldsymbol{\mu})$
```{r scores, eval=TRUE, fig.width=12, fig.height=5}
calculate_pc_scores <- function(data, pca, pc_number) {
    # Center the data
    centered_data <- scale(data, center = TRUE, scale = FALSE)
    
    # Calculate scores for specified PC
    scores <- centered_data %*% pca_rotation_clr[, pc_number]
    
    return(scores)
}

# Create data frame with scores and age
pc1_time_data <- data.frame(
    age = x_count$age,
    score = calculate_pc_scores(x_clr, pca, 1)
)

pc2_time_data <- data.frame(
    age = x_count$age,
    score = calculate_pc_scores(x_clr, pca, 2)
)

pc3_time_data <- data.frame(
    age = x_count$age,
    score = calculate_pc_scores(x_clr, pca, 3)
)
pc4_time_data <- data.frame(
    age = x_count$age,
    score = calculate_pc_scores(x_clr, pca, 4)
)


# Plot scores over time
plot1 <- ggplot(pc1_time_data, aes(x = age, y = score)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(x = "Age (ka)", 
         y = "PC1 Score",
         title = "PC1 Scores over Time") +
    theme_grey()



# Plot scores over time
plot2 <- ggplot(pc2_time_data, aes(x = age, y = score)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(x = "Age (ka)", 
         y = "PC2 Score",
         title = "PC2 Scores over Time") +
    theme_grey()



# Plot scores over time
plot3 <- ggplot(pc1_time_data, aes(x = age, y = score)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(x = "Age (ka)", 
         y = "PC3 Score",
         title = "PC3 Scores over Time") +
    theme_grey()

# Plot scores over time
plot4 <- ggplot(pc4_time_data, aes(x = age, y = score)) +
    geom_point() +
    geom_smooth(method = "loess") +
    labs(x = "Age (ka)", 
         y = "PC4 Score",
         title = "PC4 Scores over Time") +
    theme_grey()


grid.arrange(plot1, plot2, plot3, plot4, ncol = 4)    
```