---
title: "PCA on sparsely sampled densities"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: hide
---

# Introduction

Following the approach of Steyer and Greven (2023) is explained and analysed for deeper understanding of the methods at hand.
In addition, the illustrations of their paper are reproduced for practical purposes.

## Simulation

In their paper, Steyer and Greven (2023) simulate several methods:
1. Two-step approach: densities are estimated first and the PCA is conducted on the clr transformation
2. ??

### Setting

n = number of densities $f_i$ on the interval $I=[0,1]$ from a Gaussian process with true clr transformed mean function $\mu(x)=-20\left(x-\frac{1}{2}\right)^{2}+\frac{5}{3}$


### Data Generation

Generate samples for each continious density:

```{r simulate data, include = TRUE}
# Function to generate samples from the simulation setting
generate_samples_clr <- function(n = 30, m_i = 40) {
  # Define the grid points
  # x_grid <- seq(0, 1, length.out = grid_points)
  
  # Define the true clr transformed mean function
  mu <- function(x) -20 * (x - 0.5)^2 + 5/3
  
  # Define the principal component functions on clr level
  g1 <- function(x) (1/5) * sin(10 * (x - 0.5))
  g2 <- function(x) (1/10) * cos(2 * pi * (x - 0.5))
  
  # Generate factor scores for each density
  z_i1 <- rnorm(n, mean = 0, sd = sqrt(0.5))
  z_i2 <- rnorm(n, mean = 0, sd = sqrt(0.2))
  
  # Generate densities
  clr_densities <- list()
  for (i in 1:n) {
    clr_density <- function(x) mu(x) + z_i1[i] * g1(x) + z_i2[i] * g2(x)
    clr_densities[[i]] <- clr_density
  }
  
  # Sample observations from each density
  # this would be my observed dataset
  observations <- list()
  for (i in 1:n) {
    observations[[i]] <- sapply(1:m_i, function(j) {
    x <- runif(1, min = 0, max = 1)  # Zufällige Stelle x im Intervall [0, 1]
    x * clr_densities[[i]](x)
    })
  }
  
  return(list(clr_densities = clr_densities, observations = observations))
}

# inverse_clr_trafo_fc <- function(clr_composition) {
#   if (is.function(clr_composition)) {
#     return(function(x) {
#       exp(clr_composition(x) - mean(clr_composition(x)))
#     })
#   } else {
#     return(exp(clr_composition - mean(clr_composition)))
#   }
# }
```

```{r generate densities, include = TRUE}

generate_densities <- function(n = 30, m_i = 40, grid_points = 500) {
  # Define the grid points
  x_grid <- seq(0, 1, length.out = grid_points)
  
  # Define the true clr transformed mean composition
  mu <- -20 * (x_grid - 0.5)^2 + 5/3
  
  # Define the principal component compositions on clr level
  # g1 <- (1/5) * sin(10 * (x_grid - 0.5))
  # g2 <- (1/10) * cos(2 * pi * (x_grid - 0.5))
  g1 <- -(7.5/5) * sin(10 * (x_grid - 0.5))
  g2 <- (1.45) * cos(2 * pi * (x_grid - 0.5))
  
  # Generate factor scores for each composition
  z_i1 <- rnorm(n, mean = 0, sd = sqrt(0.5))
  z_i2 <- rnorm(n, mean = 0, sd = sqrt(0.2))
  
  # Generate densities
  densities <- matrix(0, nrow = n, ncol = grid_points)
  for (i in 1:n) {
    clr_density <- mu + z_i1[i] * g1 + z_i2[i] * g2
    densities[i, ] <- inverse_clr_trafo(clr_density)
    # densities[i, ] <- clrInv(clr_density)
    # densities[i, ] <- exp(clr_density)
  }
  

  
  return(densities)
}

# Function to perform inverse clr transformation
inverse_clr_trafo <- function(clr_composition) {
  exp(clr_composition) / sum(exp(clr_composition))
}

```

#### Reconstruction plots
Plot results for Figure 5 middle row:

Here the range of the pca (pca1: [-0.2, 0.2] ) is smaller than in Grevens Paper the function appears to be stretched. Both is adjustable 
but it remains unclear what the cause is. One possible explanation could be that the sampling process introduces the difference.

```{r plot  densities, include = TRUE}
# Generate samples

x_grid <- seq(0, 1, length.out = 400)
mu <- -20 * (x_grid - 0.5)^2 + 5/3
g1 <- -(7.5/5) * sin(10 * (x_grid - 0.5))
g2 <- (1.45) * cos(2 * pi * (x_grid - 0.5))

# Plot selective functions on Bayes Space
plot_data <- data.frame(x = x_grid, y = mu, pc1 = g1, pc2 = g2)

plot_data_long <- tidyr::gather(plot_data, key = "variable", value = "value", -x)

ggplot(plot_data_long, aes(x = x, y = value, color = variable)) +
  geom_line() +
  coord_cartesian(ylim = c(-2,2)) +
  labs(x = "x", y = "mu") +
  ggtitle("Plot of mu over x_grid") +
  theme_minimal() 
```


Plot densities: Figure 5 first row:

There are several differences: the variation is to low and the modes are slightly off. It is possible though that this is due to the representation. 
TODO: the scale of the density values seems off.

```{r plot densities, include = TRUE}
plot_data <- data.frame(x = x_grid, t(densities))
plot_data_long <- tidyr::gather(plot_data, key = "density", value = "value", -x)

# Create the plot using ggplot2
ggplot(plot_data_long, aes(x = x, y = value, group = density, color = density)) +
  geom_line(alpha = 0.5) +  # Adjust the alpha value for transparency
  labs(x = "x", y = "Density") +
  ggtitle("Plot of Densities") +
  theme_minimal()

# Correlation Plot
library(corrplot)
cor_matrix <- cor(densities)

# # Create the correlation plot
# corrplot(cor_matrix, method = "color", col = colorRampPalette(c("blue", "white", "red"))(100),
#          tl.col = "black", tl.srt = 45, addCoef.col = "black", 
#          title = "Correlation Plot", mar=c(0,0,2,0))
library(pheatmap)
pheatmap(cor_matrix, color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "Correlation Heatmap", angle_col = 45, 
         cluster_rows = TRUE, cluster_cols = TRUE, 
         display_numbers = FALSE, fontsize = 8)

```

## Application PCA

Firstly, we use the data generated by the 'generate_samples_simplex' function.

```{r generate samples, include = TRUE}
set.seed(123)
x_samples <- generate_samples_clr(n = 30, m_i = 40)
# we can unlist the observations to get a vector of observations
observations_clr <- as.data.frame(unlist(x_samples$observations))
observations <- exp(observations_clr)

# in Grevens setting x_data is a list of vectors
observations_clr_list <- x_samples$observations
observations_list <- lapply(observations_clr, exp)

# calculate PCA with compositional splines
spline_pca_list <- fit_compositional_spline_pca(observations_list)
```

The prcomp object contains the following components:
- sdev
- rotation
- rotated data x

## Compositional Setting

Reconstruct the same simulation as above but with descrite probability measure on a descrite Omega with 11 elements.

```{r generate compositions, include = TRUE}
# Function to generate samples from the simulation setting
generate_samples <- function(n = 30, m_i = 40, N = 11) {
  # Define the true clr transformed mean composition
  mu <- rnorm(N)
  # centering mu, s.t it sums up to zero
  mu <- mu - mean(mu)
  
  # Define the principal component compositions on clr level
  g1 <- rnorm(N)
  g1 <- g1 - mean(g1)
  g1 <- g1 / sqrt(sum(g1^2))
  
  g2 <- rnorm(N)
  g2 <- g2 - mean(g2)
  g2 <- g2 - sum(g2 * g1) * g1
  g2 <- g2 / sqrt(sum(g2^2))
  
  # Generate factor scores for each composition
  z_i1 <- rnorm(n, mean = 0, sd = sqrt(0.5))
  z_i2 <- rnorm(n, mean = 0, sd = sqrt(0.2))
  
  # Generate latent compositions
  compositions <- matrix(0, nrow = n, ncol = N)
  for (i in 1:n) {
    clr_composition <- mu + z_i1[i] * g1 + z_i2[i] * g2
    compositions[i, ] <- inverse_clr_trafo(clr_composition)
  }
  
  # Sample observations from each composition
  observations <- list()
  for (i in 1:n) {
    observations[[i]] <- sapply(1:m_i, function(j) {
      sample(1:N, size = 1, prob = compositions[i, ])
    })
  }
  
  return(list(compositions = compositions, observations = observations))
}

```

## Questions for the compositional Setting

What is the form of the "true" mean function? (the density of a multinormal distribution?)